---
title: 'Reproducible Research: Assesment 1'
output: html_document
---

For the first peer assessment, a dataset on personal movement was used. The 
collected data includes two months worth of measurements, although there are 
several missing values. The data includes the number of steps taken in 5 minute
intervals during each day.
First, we load the data into R:

```{r,tidy = TRUE}
setwd("~/Coursera/Reproducible Research/Peer assignment 1")
data <- read.csv("activity.csv", stringsAsFactors = FALSE)
data_non_NA <- data[complete.cases(data), ]
```

For the first analyses, a dataset without NA values will be used. The first 
question in the dataset revolves around the mean steps taken by day. For this,
we build a dataset that includes the total number of steps for each date.

```{r, echo=TRUE}
day_total_steps <- tapply(data_non_NA$steps, data_non_NA$date,sum)
```

We now print a histogram of all the steps per day.

```{r, tidy = TRUE}
hist(day_total_steps, main = "Histogram of steps by day", col = "sky blue",
     xlab = "Steps by day", breaks = 10)
```

We can now calculate the median (`r median(day_total_steps)`) and mean 
(`r mean(day_total_steps,digits = 4)`) of the total steps taken by day.

```{r, echo=TRUE}
summary(day_total_steps,digits = 8)
```

For the second question, we are interested know in the daily activity pattern.
We can take a look at it using the line  plot below.

```{r}
five_min_int <- tapply(data_non_NA$steps, data_non_NA$interval,mean)
plot(names(five_min_int), five_min_int, type="l", xlab = "5 min interval", 
     ylab = "Daily average", main = "Average steps per 5 minute interval",
     lwd = 2.5, col = "sky blue")
```

We can see from the chart and the calculation below that the interval with the most activity is the one that starts at 8:35 a.m., when 104 steps are taken of average.

```{r}
which.max(five_min_int)
```

There are a lot of missing values in the dataset, in fact `r sum(is.na(data$steps))`.

```{r}
sum(is.na(data$steps))
```

To try to fill them out, we will impute them using the averages of the 5 minute intervals from the non-missing dataset. We can make a histogram to see how this looks after the imputation.

```{r, tidy = TRUE}
imputed <- data
imputed[!complete.cases(imputed), "steps"] <- five_min_int[
        as.character(imputed[!complete.cases(imputed), "interval"])]
day_total_steps_imputed<- tapply(imputed$steps, imputed$date,sum)
hist(day_total_steps_imputed, main = "Histogram of steps by day", 
     col = "sky blue", xlab = "Steps by day", breaks = 10)

```

As can be seen from the table below, the summary statistics remain very similar, since the averages from the original dataset are used for the imputation.

```{r}
summary(day_total_steps_imputed)

```

Finally, we will take a look at the patterns of activity during weekends and weekdays. The chart below shows that during the weekday a lot of the activity is concentrated in the morning, while in the case of the weekends it is spread out more evenly.

```{r, tidy = TRUE}
wdays <- weekdays(as.Date(imputed$date))
wdays[wdays == "domingo" | wdays == "sábado"] <- "weekend"
wdays[wdays != "weekend"] <- "weekday"
imputed <- data.frame(imputed, wdays)

library(reshape2)
five_min_int_wday <- with(imputed,tapply(steps, list(interval,wdays),mean))
melted <- melt(five_min_int_wday,id = row.names(five_min_int_wday))
library(ggplot2)
ggplot(melted, aes(x = Var1, y = value, col = Var2)) + geom_line(size = 0.75) + 
        facet_grid(Var2~.) + xlab("5 minute interval") + 
        ylab ("Average number of steps") +
        ggtitle("Average number of steps by 5 minute interval") + 
        theme(legend.position="none")
```

